---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# autodb

autodb is an R package for automatic normalisation of a data frame to third
normal form, with the intention of easing the process of data cleaning. (Usage
to design your actual database for you is not advised.)

## Installation

You can install the development version of autodb from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("CharnelMouse/autodb")
```

## Example

Turning a simple data frame into a database:

```{r example}
library(autodb)
summary(ChickWeight)
db <- autodb(ChickWeight, name = "ChickWeight")
db
graphviz_text <- gv(db)
DiagrammeR::grViz(graphviz_text)
```

Using the `exclude_class` argument to forbid numeric variables from appearing in keys:

```{r exclude_example}
summary(CO2)
db2_noexclude <- autodb(CO2, name = "CO2")
db2_noexclude
db2 <- autodb(CO2, name = "CO2", exclude_class = "numeric")
db2
graphviz_text2 <- gv(db2)
DiagrammeR::grViz(graphviz_text2)
```

There are also functions for doing each step of the database creation separately, including functional dependency detection and normalisation. See the vignette for more details.
