---
title: "Using autodb"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using autodb}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

```{r setup}
library(autodb)
```

# Terminology

To avoid confusion, this vignette is consistent with terminology: a *database* consists of *relations*, which have *records* with the same *attributes*. This is in line with terms used in the relational model; in Statistics, where we would talk about tables, which have rows with values for the same variables, with a column for each variable. For the original data set, which is a flat table, we refer to a *data frame* instead, which also consists of records with attributes.

# Motivation

`autodb` is described in the DESCRIPTION file in the following way:

> Automatic normalisation of a data frame to third normal form, with the intention of easing the process of data cleaning. (Usage to design your actual database for you is not advised.)"

Let us begin there.

## For data cleaning

Database normalisation works on the principle that our data should only express a piece of information in a single place. This is done to reduce the chance that a modification will accidentally change or remove other pieces of information.

Data cleaning works on the principle that we remove duplicate observations and structural errors, and have a clear understanding of any associations between different observations, or parts of observations. This is done to ensure a high quality for the data used in analysis, but I have found that it is equally important for making sure that I understand the domain I am hoping to model.

Normalisation is helpful for both of these areas. I usually use the original, flat, "tidy" data for the analysis itself. However, for checking the data's consistency, and for using the data to learn about the domain, normalised data is often helpful, whether it's used for the analysis or not.

I usually do data checking under the following conditions:

- No need to store the data in normalised form. My own work normally consists of one-off analyses, where I won't see the data again once I've finished the work. For more permanent data, it would be better to store the data in a database designed to hold it, if you're able to.
- Limited documentation. In my own work, I am lucky if I am given a key or a data dictionary for the data set at all. When I am, it usually proves to be wrong somehow. This makes it even more important to do data checking, not just for looking for errors, but also as an exercise in understanding the problem domain before starting to model it.
- Data that fits in local memory. This process could be done in batches or using a streaming algorithm, but this is a more advanced case.

It's in this context that I would usually have to do the normalisation by hand, and I would like to have a tool that does it semi-automatically. This package is intended to be that tool, using a minimal amount of additional libraries.

## Not for database design

Good database design is a hard and involved process, which requires a strong grasp on the domain that the database will hold information on. Please don't try to use this package to do it on autopilot. In particular, the conditions described above mean we do things differently to good practice for normalised relational databases:

- We need to handle constant attributes, that have the same value in every record. In a real database, we would probably expect these to not stay constant when more data is added, and would know which relation to place it in. In one-off data checking, there isn't more data coming, so we need to handle these attributes in an automatic way by placing them in a "constants" relation with an empty key, which database designers would probably frown at.
- We don't care about database design principles concerned with avoiding errors under deletion, modification, and so on, because the data won't change. This means that we don't care about introducing artificial keys, unless we think it would make data modelling more convenient. Similarly, we don't care about having single-attribute reference relations, which are used to make sure other relations only assign legal values to that relation's attribute.
- We have no "derived relations", also known as "views", which are additional relations created from the others for lookup convenience. We expect to be using the data frame for modelling, rather than the database, so such lookups aren't commonly necessary, and are left to the discretion of the user to do manually with base R's `merge` function.

## To third normal form

`autodb` gets a given data frame to third normal form: every attribute depends on the whole key(s), and non-key attributes depend on nothing but the key(s). This was chosen because there is an existing algorithm, Bernstein's synthesis, for normalising to third normal form, and because it's the highest normal form that is attainable with arbitrary original data frames. There is an additional enhancement available as an option: see the section on avoidable attributes for more details.

# Simple examples

For most of these simple examples, we use the `ChickWeight` dataset, included with base R.

```{r}
summary(ChickWeight)
```

## Normalisation with `autodb`

The simplest way to do normalisation is with the `autodb` function:

```{r}
db <- autodb(ChickWeight, name = "Chick Weights")
db
```

## Plots

There is no plotting method within the `autodb` package itself. Instead, there are functions to write normalised objects as inputs to the Graphviz visualisation software, using the DOT language.

The generic function to do this is `gv`, which calls the `gv.database` method on our database, `db`:

```{r}
db_text <- gv(db)
cat(db_text)
```

This can be saved to a file for passing to Graphviz elsewhere, or can be passed into `grViz` in the `DiagrammeR` package:

```{r db_plot}
DiagrammeR::grViz(db_text)
```

Each relation is represented as a box, with the top row giving the relation name and the number of records, and the other rows detailing the attributes. In the middle is a grid of cells: each column of cells represents a (candidate) key for the relation, where attributes in that key have their cell filled in black.

We can see that `ChickWeight` was split into two relations, with a weight for each unique combination of Chick number and time, and the Chick number by itself determining the diet. We can also see the number of records in each relation, and the classes of the attributes in those relations.

`gv` also has methods for `data.frame`, and for `database_schema`, which is introduced later.

## Finding functional dependencies

Having looked at the final result first, we now look at the individual steps. The first step is to find the (non-trivial and minimal) dependencies present in the given data frame. There are various ways to do this; by default, the package uses DFD, a depth-first search algorithm. We run this using the `discover` function, setting `progress` to `TRUE` to see the steps taken:

```{r}
deps <- discover(ChickWeight, accuracy = 1, progress = TRUE)
```

The `accuracy` argument expects a number between zero and one, determining what proportion of a data frame's records need to satisfy a given dependency for the algorithm to consider it valid. Setting `accuracy` to one, as done here, limits us to exact dependencies. We'll look at other values later, in the section on approximate dependencies.

```{r}
deps
```

After simplifying the data to something quicker to iterate over -- specifically, by converting all attributes to have integer values -- the algorithm takes each attribute in turn as a possible dependent, and finds sets of the other attributes that determine it.

The result is a list of functional dependencies, in the format `determinant set -> dependent`, with an attribute named `attrs_order` that gives all the attribute names in their original order. Each of these three parts has its own (generic) extraction function:

```{r}
detset(deps)
dependent(deps)
attrs_order(deps)
```

The former two, especially, are useful for filtering, as we'll see later.

## Normalisation

Now that we have a list of discovered dependencies, we can construct a database schema, where the relation schemas are normalised to third normal form. This is done using a version of Bernstein's synthesis.

```{r}
schema <- synthesise(deps)
schema
```

Like the database before, we can also plot this database schema:

```{r}
schema_text <- gv(schema)
cat(schema_text)
DiagrammeR::grViz(schema_text)
```

This is similar to the database plot given before, but there is some information not present, that requires the data frame itself. Namely, we have no class information about the attributes, and no record counts for the individual relation schemas. Additionally, at this point we have no connections between the relation schemas, since Bernstein's synthesis doesn't supply information about foreign key references. We could use this database schema to build a database, but we'd rather add the foreign key references first.

## Tuning detection and normalisation

Let's look at a different dataset for a moment, to look at some cases where we don't want to use the dependencies as given. We'll use the Titanic data set, also provided with base R. This data is in array form, so we first convert it to data frame form:

```{r}
as.data.frame(Titanic)
```

This is a simple set of data, with a single count observation, `Freq`, for each combination of the four determining attributes. In other words, the relation is already normalised, so we only expect one relation in the normalised database.

If we use `autodb` again, we get the following database:

```{r}
DiagrammeR::grViz(gv(autodb(as.data.frame(Titanic))))
```

Oops! The DFD algorithm found some functional dependencies where the count could be used to determine another attribute. These are clearly spurious: frequency count can't causally determine age, for example. However, the algorithm still finds these dependencies, because the counts are unique often enough to make these dependencies hold in the given data.

There are two approaches we can take to eliminate these spurious dependencies: not letting them be detected in the first place, and removing them before using `synthesise`.

To stop them being detected, we can put constraints on what is discovered by `discover`: we can ask for certain attributes to not be considered as determinants, or we can exclude attributes that inherit from certain classes. In this example, we could exclude `Freq` from being considered:

```{r}
titanic_deps_freqonly <- discover(as.data.frame(Titanic), 1, exclude = "Freq")
titanic_deps_freqonly
```

Alternatively, we could exclude all attributes that inherit from "numeric":

```{r}
identical(titanic_deps_freqonly, discover(as.data.frame(Titanic), 1, exclude_class = "numeric"))
```

These can both be used as arguments to `autodb` too:

```{r}
DiagrammeR::grViz(gv(autodb(as.data.frame(Titanic), exclude = "Freq")))
```

Generally, excluding numeric attributes as determinants is often useful, because we expect non-integer numbers to be a measurement, not part of a primary key.

Alternatively, we could remove unwanted dependencies before using `decompose`. Here are the found dependencies, if we don't exclude anything:

```{r}
titanic_deps <- discover(as.data.frame(Titanic), 1)
titanic_deps
```

We can remove the unwanted dependencies, where `Age` is the dependent:

```{r}
titanic_deps[dependent(titanic_deps) == "Freq"]
```

## Adding cross-references

Getting back to our `ChickWeight` example, we now have a database schema, consisting of a list of `relation` schemas. However, we have no information about how these relation schemas are linked to each other. In particular, we have no information about foreign keys. We can add this information using `cross_reference`:

```{r}
linked_schema <- cross_reference(schema)
linked_schema
```

We could also have used `normalise`, instead of `synthesise` and `cross_reference` separately:

```{r}
normalise(deps)
```

Plotting this updated database schema shows the same relation schemas as before, linked together by foreign key references:

```{r}
linked_text <- gv(linked_schema)
cat(linked_text)
DiagrammeR::grViz(linked_text)
```

## Decomposing the original relation

Finally, once we have our normalised database schema, we can apply it to our original data frame, or a new one with the same structure. This results in a normalised database, as we got from using `autodb`:

```{r chickWeight_db2_plot}
db2 <- decompose(ChickWeight, linked_schema)
DiagrammeR::grViz(gv(db2, name = "Chick Weights"))
```

We now have the attribute class information and record counts added, as well as automatic names for the individual relations.

## Rejoin a database to a data frame

We can reverse the process of turning a data frame into a database with the `rejoin` function. This may not be identical to `ChickWeight`, since the rows may have been rearranged. However, we can use the convenience function `df_equiv` to check for equivalence under row reordering:

```{r chickWeights_rejoin}
rejoined <- rejoin(db)
summary(rejoined)
identical(rejoined, ChickWeight)
df_equiv(rejoined, ChickWeight)
```

When rejoined, the relation attributes will be in the original order. However, the record order might have changed.

# A larger example

Included in the package is a `r paste(dim(nudge), collapse = "-by-")` data frame called `nudge`:

```{r nudge_classes}
DiagrammeR::grViz(gv(nudge, name = "nudge"))
```

This is the data set for a meta-analysis, looking at the effectiveness of "nudge" interventions. Measurements are taken within a three-layer hierarchy: publications contain studies, which contain effect size measurements. We expect this hierarchy to be apparent in the normalisation.

Getting full dependency information for this relation can take a long time, so here we search for a reduced set, not considering numeric or sample size attributes as determinants:

```{r nudge_database}
nudge_deps <- discover(
  nudge,
  accuracy = 1,
  exclude = c("n_study", "n_comparison", "n_control", "n_intervention"),
  exclude_class = "numeric"
)
nudge_schema <- normalise(nudge_deps, remove_avoidable = TRUE)
DiagrammeR::grViz(gv(nudge_schema))
```

We can see a relation, with many attributes, determined by the effect size ID number, `es_id`. This contains all of the numeric measurements, as expected in the relation for the lowest level in the hierarchy. As also expected, this has a foreign key reference to a relation for study-level information, determined by the study ID number, `study_id`.

However, the publication-level relation this refers to is not determined by the publication ID, `publication_id`, as we would expect; neither is it determined by the publication reference. Instead, it is solely determined by the publication's title: to use the ID, we would need to supplement it with the publication year. This suggests that some publication ID numbers have been erroneously assigned to several publications, which we can easily test:

```{r nudge_publication_check}
nudge_database <- decompose(nudge, nudge_schema, "nudge")
nudge_title_relation <- nudge_database$relations$title$df
nudge_pid_duplicates <- unique(nudge_title_relation$publication_id[
  duplicated(nudge_title_relation$publication_id)
])
knitr::kable(subset(nudge_title_relation, publication_id %in% nudge_pid_duplicates))
```

This would also explain why the publication ID appears in several other relations, that look likely to be spurious.

Since the publication ID can't be trusted, we could consider removing the attribute, and re-running `discover`. While fine in this case, this is best avoided if running `discover` on a given data frame is expensive.

Another option is to simply remove the offending relations in the current schemas:

```{r nudge_publication_badfilter}
nudge_schema_relfiltered <- nudge_schema[!grepl("publication_id", names(nudge_schema), fixed = TRUE)]
```

Subsetting a database schema also removes any foreign key relationships involving removed schemas, so the resulting schema is still valid. However, any foreign key reference chains with removed schemas in the middle are broken. Amongst others, in this case we lose the relationship between the three relations for the three hierarchical data levels:

```{r nudge_publication_badfilter_gv}
DiagrammeR::grViz(gv(nudge_schema_relfiltered))
```

We also lose any functional dependencies partially implied by the removed schemas (i.e. transitive dependencies, where removing these schemas' dependencies means that they're no longer transitive). This is an issue if those partially-implied dependencies aren't spurious, because we throw away non-spurious information about the data's structure.

The better option is to remove all functional dependencies where the publication ID is in the determinant set, to get the same effect without re-running `discover`:

```{r nudge_filter}
nudge_deps_filtered <- nudge_deps[!vapply(detset(nudge_deps), is.element, logical(1), el = "publication_id")]
nudge_schema_filtered <- normalise(nudge_deps_filtered, remove_avoidable = TRUE)
DiagrammeR::grViz(gv(nudge_schema_filtered))
```

The resulting database schema looks to be an improvement. However, we can still see some spurious relations involving `reference`. One of these combines it with `type_experiment` to determine `location`, a study-level attribute. The others combine it with another attribute to determine `title`.

As mentioned, we'd expect the reference to uniquely identify the publication: however, in the publication-level relation, the title determines the reference, but not vice versa. This means that there are publications that share a reference:

```{r nudge_reference_check}
nudge_database_filtered <- decompose(nudge, nudge_schema_filtered, "nudge")
nudge_title_relation_filtered <- nudge_database_filtered$relations$title$df
nudge_reference_duplicates <- unique(nudge_title_relation_filtered$reference[
  duplicated(nudge_title_relation_filtered$reference)
])
knitr::kable(subset(nudge_title_relation_filtered, reference %in% nudge_reference_duplicates))
```

BETA is the Behavioural Economics Team of the Australian Government, so it's not surprising that they'd have multiple publications/reports per year. Duplicate references is not necessarily an error, but would be awkward if the references were to be used.

While the reference is useful, the relations mentioned above are not, so we'll also remove functional dependencies with the reference determining something apart from the year:

```{r nudge_filter2}
nudge_deps_filtered2 <- nudge_deps_filtered[
  !vapply(detset(nudge_deps_filtered), is.element, logical(1), el = "reference") |
    dependent(nudge_deps_filtered) == "year"
]
nudge_schema_filtered2 <- normalise(nudge_deps_filtered2, remove_avoidable = TRUE)
DiagrammeR::grViz(gv(nudge_schema_filtered2))
```

We now, finally, have the study location in a non-spurious relation. Interestingly, removing dependencies has revealed extra information about study locations, via the new `title_type_experiment` relation: studies of the same experiment type in a publication always have the same location. Looking at the resulting database shows that this removes many entries of what would be redundant location information if kept in the study relation:

```{r nudge_clean_database}
nudge_database_filtered2 <- decompose(nudge, nudge_schema_filtered2, "nudge")
DiagrammeR::grViz(gv(nudge_database_filtered2))
```

While this is not a dependency we could expect to hold if more data was collected, it's a reasonable dependency for the given data set, which won't be added to.

It's also notable for the previous discussion about information lost if we subset the database schema: the relation isn't present in the original schema because the full set of functional dependencies make its implied dependency transitive, but removing spurious dependencies made it no longer transitive, so it's directly represented in a new relation. Specifically, the previous schema, `nudge_schema_filtered`, has relations showing that `title -> reference` and `reference, type_experiment -> location`, and removing the latter makes `title, type_experiment -> location` no longer transitive. Subsetting the schema, instead of removing the spurious dependencies, would also lose this transitive dependency, so we would not see the new relation.

(As far as I'm aware, the publication ID error mentioned above would not have affected the meta-analysis results.)

# Other features

## Approximate dependencies and database reduction

Larger datasets can often have entry errors, without an easy way to remove or deal with them. For this reason, we might be interested in "approximate" functional dependencies, which hold after removing a bounded amount of violating records.

Suppose that we normalise `nudge` again, without any manual dependency removal, but allowing approximate dependencies. We could cheat, knowing that the questionable data example found above showed there to be two questionable records: one for a duplicated publication ID, and one for a duplicated reference. Since `nudge` has `r nrow(nudge)` records, we can get rid of the resulting questionable dependencies by setting the accuracy to allow two discrepant records.

```{r nudge_approximate_cheat}
nudge_approx_cheat_database_schema <- discover(
  nudge,
  accuracy = 1 - 2/nrow(nudge),
  exclude = c("n_study", "n_comparison", "n_control", "n_intervention"),
  exclude_class = "numeric"
) |>
  normalise()
DiagrammeR::grViz(gv(nudge_approx_cheat_database_schema))
```

Currently `decompose` doesn't account for approximate dependencies, resulting in invalid databases, so here we just work with the database schema.

Compare this to the database schema we arrived at manually:

- The publication ID, the title, and the reference are considered equivalent, as we'd have originally expected.
- There is an additional relation, showing `approximation` to be approximately determined by some publication- and study-level attributes, rather than functionally determined at the effect-size level.

Lowering the accuracy results in more dependencies being found, and so in more relations. The number of relations can get very large. For example, suppose we instead set the accuracy to `0.99`, returning any approximate dependencies that hold in at least `r ceiling(nrow(nudge)*0.99)` records.

```{r nudge_approximate}
nudge_approx_database_schema <- discover(
  nudge,
  accuracy = 0.99,
  exclude = c("n_study", "n_comparison", "n_control", "n_intervention"),
  exclude_class = "numeric"
) |>
  normalise()
DiagrammeR::grViz(gv(nudge_approx_database_schema))
```

This is a little overwhelming, so we'll use a utility function called `reduce`. This finds the relation with the most records, which can be considered as the "main" relation, and removes any relations not linked to it by foreign key references. Reducing the approximate database gives us this set of relations:

```{r nudge_approximate_reduced}
DiagrammeR::grViz(gv(reduce(nudge_approx_database_schema, "es_id")))
```

Questionable intermediate relations aside, we can see that there is now a publication-level relation with the publication ID as a key, since it determines the publication attributes once we remove one of the duplicate records we discovered before.

`reduce` can also be used on databases, where it considers each relation with the largest number of records as a main table. It needs to be used with caution: while the relations it removes are often spurious, it's possible to find databases where relations not linked to the main relation by foreign key references are required to rejoin to the original data frame, so reduction can remove necessary relations. Its intent is mostly to make glancing at database plots more manageable.

## Avoidable attributes

The next normal form after third normal form (3NF) is Boyes-Codd normal form (BCNF). It represents an issue for automatic normalisation, since not all relations can be put into BCNF. Even worse, determining whether a database schema in 3NF is in BCNF is intractable.

Instead, the package includes an option to convert to enhanced third normal form, also known as LTK form. This enhancement is tangential to BCNF, and can also be used to enhance schemas in BCNF.

In brief, the standard normal forms only put constraints on the attributes present in the relations one relation at a time. The enhancement is a constraint on the attributes present in a relation, while considering their presence in other relations. If a attribute in a relation can be removed, and still be determined from that relation by joining it to others, then the attribute is "avoidable", and can be removed. If the attribute is in any of the relation's keys, they'll be replaced by keys that use the attributes not being removed. This removes attributes from relations without removing any information from the database as a whole.

For example, we can take this simple example from Chapter 6 of The Theory of Relational Databases, by David Maier:

```{r avoid_setup}
avoid_deps <- functional_dependency(
  list(
    list("A", "B"),
    list("B", "A"),
    list(c("A", "C"), "D"),
    list(c("A", "C"), "E"),
    list(c("B", "D"), "C")
  ),
  attrs = c("A", "B", "C", "D", "E")
)
avoid_deps
avoid_schema <- normalise(avoid_deps)
DiagrammeR::grViz(gv(avoid_schema))
```

Attributes `A` and `B` are equivalent, since relation `A` has them both as a key. In other words, relation `A` is a simple lookup relation. Because of this, we could remove `B` from relation `A_C`, and replace the key `B, D` with `A, D`, which is equivalent when accounting from the lookup relation `A`.

We can have this removal of avoidable attributes done automatically, using the `remove_avoidable` flag for `normalise`:

```{r avoid_remove}
avoid_schema_removed <- normalise(
  avoid_deps,
  remove_avoidable = TRUE
)
DiagrammeR::grViz(gv(avoid_schema_removed))
```

This schema is now in LTK form, with no remaining avoidable attributes. We could have also removed `A` from relation `A_C` instead of `B`, so this process may not have a unique result. The package's implementation prefers to remove attributes that appear later in the original relation.

# Planned extensions

## Class information in schemas

As shown in the plots above, database schemas don't contain information about the attribute classes: this is only added on database creation. This would potentially be useful if the original data frame is large, and creating the database would take a long time.

## Handling of duplicate records

It's expected that an R user might pass a data frame into the package that has duplicate records. At the moment, these are kept when searching for dependencies. This can affect results for approximate dependencies, due to affecting the record counts. However, they will be removed when the data frame is decomposed into a database. At the moment, I'm not certain on whether they are best handled by removing them, or by simply returning an error if there are duplicate records.

## Missing values

In relational databases, we commonly want to avoid the presence of missing values. For example, take the following data frame:

```{r example_data_frame_with_NAs}
df_nas <- data.frame(
  patient_id = c(1L, 2L, 3L, 4L),
  trial_entry_date = as.Date(c("2022/05/02", "2022/06/06", "2022/04/01", "2022/03/19")),
  death_date = as.Date(c(NA, NA, "2022/10/07", NA))
)
df_nas
```

The package currently treats `NA` as just another value, with no special treatment for normalisation or rejoining, resulting in no splitting of the above data frame:

```{r example_data_frame_with_NAs_autodb}
DiagrammeR::grViz(gv(autodb(df_nas)))
```

In this case, a missing death date represents no death, and we would prefer to move death information to a separate relation, containing only patients with a death date, in this case patient 3:

```{r example_data_frame_with_NAs_nullably_normalised}
ideal_db <- decompose(
  df_nas,
  database_schema(
    relation_schema(
      list(
        patient = list(c("patient_id", "trial_entry_date"), list("patient_id")),
        death = list(c("patient_id", "death_date"), list("patient_id"))
      ),
      names(df_nas)
    ),
    list(list(2:1, c("patient_id", "patient_id")))
  )
)
ideal_db$relations$death$df <- subset(ideal_db$relations$death$df, !is.na(death_date))
DiagrammeR::grViz(gv(ideal_db))
```

Handling cases like this requires automatically detecting, and normalising, nullable functional dependencies, which account for the ambiguity of missing values.
